Kindly Bent to Free Us

### 2019-07-06 kindly/infer/infer.tex
Start working on the inference section.

\section{Type inference}

We now formulate type inference for the \lang language. Type inference is
based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
presents how to implement Hindley-Milner type inference in a language
with qualified types and an arbitrary custom constraint language.
We first present various preliminaries definitions,
we then show our extension of HM(X) that support a kind system and affine
types.

@@

\subsubsection{Usage maps}

In order to determine if a variable is used in an affine manner, we must track
its uses and the associated kinds. For instance, in the expression
$(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
we must add the constraint $\Cleq{k}{\kun}$. In order to record uses and kinds,
our type inference relation will return a ``usage map'', noted $\Sv$,
which associates each variable to a list of kinds corresponding to each use site
of the variable.
We note $\Sone{x}{k_1,\dots,k_n}$ the singleton usage map.
Common operations on usage maps are defined in \cref{op:usgmap}. We define
regular set operations such as union
($\Sv\cup\Sv'$) and intersection ($\Sv\cap\Sv'$), along
with constraint creation $\Cleq{\Sv}{k_0}$ which indicates that all the kinds
presents in the usage map $\Sv$ are constrained by the kind $k_0$.
These operations corresponds roughly to the various parts of the join
operator on sets $\esplit{}{}$.

\begin{figure}[tp]\centering
\begin{mathpar}
  % \Sv\cap\Sv'
  % &= \left\{ x \to l,l'
  %   \mid \Sone{x}{l}\in\Sv \text{ and } \Sone{x}{l'}\in\Sv' \right\} \\
  % \infer
  % { \forall x \in \Dom{\Sv_1}\cup\Dom{\Sv_2}.\quad
  %   \bsplit{C_x}{\Sv(x)}{\Sv_1(x)}{\Sv_2(x)}
  % } 
  % { \bsplit{(\bigwedge C_x)}{\Sv}{\Sv_1}{\Sv_2} }
  % &= \Sv\cap\Sv' \sqcup
  %   (\Sv\Sdel{x' \mid x'\in\Sv'}) \sqcup
  %   (\Sv'\Sdel{x \mid x\in\Sv})
  %   &\text{where $\sqcup$ is the disjoint union}\\
  % \and
  % \inferrule{}
  % { \Cleq{\Sv}{k}
  %   = \bigwedge_{\Sone{x}{\schm}\in\Sv} \Cleq{\schm}{k}
  % }
  % \and
  % \inferrule
  % { \forall \bshadow{x} \in \Sv, b'\in\Sv'.\quad
  %   \tregion{C_x}{y}{\bshadow{x}}{b'} }
  % { \bregion{(\bigwedge C_x)}{y}{\Sv}{\Sv'} }
  % \and
  % \inferrule
  % { \text{if } \operatorname{lenght}(\Sv(x))  = \Cempty{} \wedge  }
  { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
    \begin{dcases}
      \Ctrue& \text{if } x\in\Sv\\
      \Cleq{\sigma}{\kaff_\infty}&\\
    \end{dcases}

  }

@@

We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
in environment $\E$ under constraints $C$ and unifier $\unif$. From an
algorithmic point of view, $\E$ and $\tau$ are the input parameters of
our inference procedure.
We present the kind inference algorithm as a set of syntax-directed rules in
\cref{rules:kinding}.
Similar to the syntax-directed rules, higher-kinds are not generally supported
and can only appear by looking-up the kind scheme of a type constructor,
for use in the type application rule {\sc KApp}.
Type variables must be of a simple kind in rule {\sc KVar}.
Kind schemes are instantiated in the {\sc KVar} rules by creating
fresh kind variables and the associated substitution.
{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
arrow and borrow types.

\begin{figure}[ht]
  \centering
  \input{infer/kinds}
  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{(\Multi{k}) k}$}
  \label{rules:kinding}
\end{figure}

We reformulate the HM(X) type inference in the context of our affine type
system. The main difference compared to HM(X) are noted in \addlin{blue}.
We note $\inferW{\addlin{\Sigma}}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$.
$\Sigma$ is a map which associates free variables in $e$ to
their kinds.
As before, $\E$ and $e$ are the input parameters of the inference
algorithm. The syntax-directed rules are shown in \cref{rules:typing}.

\subsection{Normal forms and generalization}
\label{sec:normalize}

@@

\subsubsection{Soundness and Principality}

@@

\subsection{Principality}

>>>>

\section{Inference}

We now formulate type inference for \lang. Type inference is
based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
presents how to implement Hindley-Milner type inference in a language
with qualified types and an arbitrary custom constraint language.
We adapt and extend the rules of HM(X) to support kind inference,
track linearity and affinity and handle borrow and regions. Furthermore, we
formulate an appropriate constraint solving algorithm for our constraint
system such that inference is sound and complete.
In the rest of this section, we first present various preliminaries definitions
before showing our kind and type inference judgements
and our constraint solving algorithm.

@@

Unlike the syntax-directed version of our type system, knowing which elements
are input and output or our inference judgement is critical. In the rest
of this section, when presenting a new judgement
we will note in \textbf{bold} the input parameters. The rest will then be
output parameters.

\subsubsection{Usage environments}

% One novelty of our inference  judgement now
% returns a type environment, which we call ``usage environment'' and commonly
% note $\Sv$, to summarize how variables and borrows are used inside
% the expression.

In order to determine if a variable is used in an affine manner, we must track
its uses and the associated kinds. For instance, in the expression
$(x,x)$, $x$ is used twice. If $x$ is of type $\tau$, which is of kind $k$,
we must add the constraint $\Cleq{k}{\kun}$.
%
In order to infer such constraints, our inference judgement will not only
take an environment as parameter but also return an environment, which
we call ``usage environment'', and which summarize how variables and borrows
are used. Usage environments follow the exact same grammar
as normal environment. In order to distinguish them more easily,
we will general note them $\Sv$.

In \cref{sdtyping}, we used various relations to split environments in two, or
transform suspended bindings into borrow binding inside an environment.
These relations took as argument a constraint which is used to validate
the environment transformation.
In the context of inference, we define a new judgements which \emph{infers}
the constraints required.
\begin{itemize}
\item $\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$.
  Given two usage environments (inferred for two subexpressions, for instance)
  $\Sv_1$ and $\Sv_2$, we return $\Sv$, the merged environment, and $C$, a set
  of constraints that must be respected.
  This relation is total and non-ambiguous given $\Sv_1$ and $\Sv_2$
  and uses the same rules as the one presented in \cref{sdtyping:split}.
\item $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$.
  Given a usage environment $\Sv'$ and a variable name $x$, we return
  $\Sv$ where the borrow binding of $x$ in $\Sv'$, if it exists, is replaced by
  a suspended binding. We also return $C$, a set of constraints that must
  be respected.
  Again, the relation is total and non-ambiguous for any given $\Sv'$ and $x$,
  and uses the same rules as the one presented in \cref{sdtyping:regions}.
\end{itemize}

The relations used for syntax-directed typing can trivially be defined
in term of these new relations and constraint entailment.
All relations are fully described in \cref{typ:extra:envs}.

\subsubsection{Constraint normalization}

The HM(X) framework assumes the existence of a $\operatorname{normalize}$
function which takes a constraint $C$ and a substitution $\psi$ and returns a
constraint in solved form $C'$,
and an updated substitution $\unif'$.
Normalization is supposed to return the ``best'' solution $(C',\unif')$, for
instance the most general unifier.
We detail the implementation
of the normalization function in \cref{infer:solving}. In the meantime, we just
assume the existence of such a function for our constraint system.

We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
in environment $\E$ under constraints $C$ and unifier $\unif$.
$\E$ and $\tau$ are the input parameters of
our inference procedure.
We present the kind inference algorithm as a set of rules in
\cref{rules:kinding}.
Similar to the syntax-directed rules, higher-kinds are not generally supported
and can only appear by looking-up the kind scheme of a type constructor,
for use in the type application rule {\sc KApp}.
Type variables must be of a simple kind in rule {\sc KVar}.
Kind schemes are instantiated in the {\sc KVar} rules by creating
fresh kind variables and the associated substitution.
{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
arrow and borrow types.
The $\operatorname{normalize}$ function is used every time several constraints
must be composed in order to simplify the constraint and return a most general
unifier.

\begin{figure}[ht]
  \centering
  \input{infer/kinds}
  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
  \label{rules:kinding}
\end{figure}

We note $\inferW{\Sv}{(C,\unif)}{\bf{\E}}{\bf{e}}{\tau}$ when
$e$\ has type $\tau$ in $\E$ under the constraints $C$ and unifier $\unif$
with a usage environment $\Sv$. $\E$ and $e$ are the input parameters of our
inference algorithm.
The rules are shown in \cref{rules:typing}.

\TODO{Explain/Cut}

\begin{figure}[tp]
  \centering
  \begin{mathpar}
    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
      \begin{dcases}
        \Ctrue& \text{if } x\in\Sv\\
        \Cleq{\sigma}{\kaff_\infty}&\\
      \end{dcases}
    }

@@

\subsection{Constraint solving}
\label{infer:normalize}

\TODO{Present the full constraint system}

@@

\subsection{Soundness and Principality}

\TODO{Simplify/Cut}

@@
